---
title: "Motor vehicle ownership in DC"
author: "pete rodrigue"
date: "2026-02-18"
output: html_document
---

```{r setup, include=F, echo=F, warning=F, message=F}
knitr::opts_chunk$set(echo=F, warning=F, message=F)
# remove.packages("ggplot2")
# install.packages("ggplot2")
library(readr)
library(sf)
library(dplyr)
library(leaflet)
library(leaflet.extras)
library(rstudioapi)
library(ggplot2)
# devtools::install_github("liamgilbey/ggwaffle")
library(ggwaffle)
library(patchwork)
library(tigris)
library(maptiles)    # fetch tile basemap as a SpatRaster
library(tidyterra)   # geom_spatraster_rgb() to plot it in ggplot2
library(ggspatial)   # north arrow + scale bar


# set working directory to be the folder that this Rmd file is saved in, 
# for convenience 
current_path = rstudioapi::getActiveDocumentContext()$path 
setwd(dirname(current_path))
# print(getwd())

# load car ownership data
co <- readr::read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSOGUHkvJwIW_vFfjCPjvBQBW2K9JAfl7An78DTK5CLW-LIrFRBxBMgexXweqZbgLpOUMZXxwbwyF30/pub?gid=1380852961&single=true&output=csv") %>%
  sf::st_as_sf(., coords = c("Geocodio Longitude", "Geocodio Latitude"), crs = 4326)

# https://opendata.dc.gov/datasets/DCGIS::single-member-district-from-2023/about
smd <- sf::st_read("data/Single_Member_District_from_2023/Single_Member_District_from_2023.shp") %>%
  sf::st_transform(crs = 4326)

# https://opendata.dc.gov/datasets/DCGIS::advisory-neighborhood-commissions-from-2023/about
anc <- sf::st_read("data/Advisory_Neighborhood_Commissions_from_2023/Advisory_Neighborhood_Commissions_from_2023.shp") %>%
  sf::st_transform(crs = 4326)

# https://opendata.dc.gov/datasets/DCGIS::address-points/about
pts <- readr::read_csv("data/Address_Points.csv") %>%
  sf::st_as_sf(., coords = c("LONGITUDE", "LATITUDE"), crs = 4326)

# https://opendata.dc.gov/datasets/a53c0f02804a484b87027ce3ef3ff38b_41/about
trt_d <- sf::st_read("data/ACS_5-Year_Economic_Characteristics_DC_Census_Tract/ACS_5-Year_Economic_Characteristics_DC_Census_Tract.shp") %>%
  select(TRACTCE, DP03_0062E) %>%
  rename(
         med_hh_inc = DP03_0062E,
         TRACT = TRACTCE
         ) %>%
  sf::st_transform(crs = 4326)

# https://opendata.dc.gov/datasets/DCGIS::census-tracts-in-2020/about
trt <- sf::st_read("data/Census_Tracts_in_2020/Census_Tracts_in_2020.shp") %>%
  select(TRACT, H0010001, P0030001) %>%
  rename(
         num_hh = H0010001, 
         pop_over_18 = P0030001
         ) %>%
  sf::st_transform(crs = 4326)

# https://opendata.dc.gov/datasets/DCGIS::metro-stations-regional/about
mto <- 
  sf::st_read("data/Metro_Stations_(Regional)/Metro_Stations_(Regional).shp") %>%
  mutate(state = substr(ADDRESS, nchar(ADDRESS) - 2 + 1, nchar(ADDRESS))) %>%
  filter(state == "DC")

# https://opendata.dc.gov/datasets/DCGIS::alleys-and-parking/explore?location=38.941233%2C-77.079861%2C15
pk <- sf::read_sf("Alleys_and_Parking/Alleys_and_Parking.shp", quiet = T) %>%
  filter(DESCRIPTIO == "Parking Lot")

drop_units <- function(x) {
  units(x) <- NULL
  
  return(x)
}

```


In January 2026, there were `r format(sum(co[["Count of Vehicles"]]), big.mark = ",")` vehicles registered with the DC DMV at `r format(length(unique(co$Address)), big.mark=",")` residential addresses, according to data provided in a recent FOIA request.

Keep in mind that this does not include any vehicles that are owned by DC residents, but registered in another state. This also doesn't include vehicles registered at non-residential addresses, like rental cars, university vehicles, or federally-owned vehicles. 

The DC DMV website says there are [310,000](https://dmv.dc.gov/page/about-dc-dmv) total vehicles registered in DC. 

According to a 2023 U.S. DoT survey, there are [344,088](https://www.fhwa.dot.gov/policyinformation/statistics/2023/mv1.cfm) total motor vehicles registered in DC, of which [315,319](https://www.fhwa.dot.gov/policyinformation/statistics/2023/mv1.cfm) are privately owned (i.e. not owned by a government agency), which roughly aligns with the DMV website's 310k figure. 

If we add up the number of 7.5'x21' parking spaces required for these 315,319 vehicles, we get a total area of 1.79 square miles (158 sqft * 315,319 vehicles * 3.58701e-8 sqmi/sqft). If you parked all of these vehicles bumper to bumper, they would cover Capitol Hill. 

All 344,088 vehicles would cover closer to 2 square miles. 

According to the 2020-2024 American Community Survey, [36%](https://data.census.gov/table/ACSDT5Y2024.B08201?t=Transportation&g=040XX00US11&y=2024) of DC households do not have a motor vehicle.

```{r static_map}
options(tigris_use_cache = TRUE)

# ── Helper: build square polygon ─────────────────────────────────────────────
make_polygon <- function(center_lat, center_lon, sqmi) {
  target_area_m2 <- sqmi * 2589988.11
  side_m         <- sqrt(target_area_m2)
  lat_deg        <- side_m / 111320
  lon_deg        <- side_m / (111320 * cos(center_lat * pi / 180))
  half_lat       <- lat_deg / 2
  half_lon       <- lon_deg / 2
  
  polygon_coords <- matrix(
    c(
      center_lon - half_lon, center_lat - half_lat,
      center_lon + half_lon, center_lat - half_lat,
      center_lon + half_lon, center_lat + half_lat,
      center_lon - half_lon, center_lat + half_lat,
      center_lon - half_lon, center_lat - half_lat
    ),
    ncol = 2, byrow = TRUE
  )
  
  st_sf(geometry = st_sfc(st_polygon(list(polygon_coords)), crs = 4326))
}

# ── Data ──────────────────────────────────────────────────────────────────────
center_lat <- 38.8897
center_lon <- -76.993

private_vehicles_poly <- make_polygon(center_lat, center_lon, sqmi = 1.79)

# DC boundary from TIGRIS (state FIPS 11 = District of Columbia)
dc_boundary <- states(cb = TRUE, progress_bar = FALSE) |>
  filter(STATEFP == "11") |>
  st_transform(4326)

# ── Fetch basemap tiles clipped to DC extent ──────────────────────────────────
# maptiles fetches tiles for a given sf object's bounding box
basemap_tiles <- get_tiles(
  x        = dc_boundary,
  provider = "CartoDB.Positron",
  zoom     = 12,
  crop     = TRUE   # crop raster to the exact dc_boundary bbox
)

# Mask tiles to DC boundary shape (sets pixels outside DC to NA)
basemap_masked <- terra::mask(basemap_tiles, terra::vect(dc_boundary))

# ── Map ───────────────────────────────────────────────────────────────────────
p <- ggplot() +
  
  # Basemap raster, masked to DC boundary
  geom_spatraster_rgb(data = basemap_masked) +
  
  # DC boundary outline
  geom_sf(data      = dc_boundary,
          fill      = NA,
          color     = "grey30",
          linewidth = 0.6) +
  
  # Capitol Hill square polygon
  geom_sf(data      = private_vehicles_poly,
          fill      = "#B22222",
          color     = "#B22222",
          alpha     = 0.25,
          linewidth = 0.8) +
  
  # Crop plot to DC bounding box
  coord_sf(
    xlim   = st_bbox(dc_boundary)[c("xmin", "xmax")],
    ylim   = st_bbox(dc_boundary)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  
  # North arrow + scale bar
  # annotation_north_arrow(
  #   location    = "tr",
  #   which_north = "true",
  #   style       = north_arrow_fancy_orienteering(text_size = 8)
  # ) +
  # annotation_scale(location = "bl", width_hint = 0.3) +
  
  labs(
    # title    = "Capitol Hill — Private Vehicles Coverage (~1.79 sq mi)",
    # subtitle = "Washington, D.C.",
    caption  = "Basemap © CartoDB / OpenStreetMap contributors"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "grey40", size = 10),
    plot.caption  = element_text(color = "grey60", size = 7,
                                 margin = margin(t = 6)),
    plot.margin   = margin(10, 10, 10, 10)
  )

p
```


This is a massive underestimate of the amount of space actually taken up by parking, however. To get a rough, likely low-ball estimate of the amount of parking in DC, you can add up the following:

* the area of DC's parking lots (5.14 sqmi)
* the area of DC's ~150,000 RPP (`r round(150000 * 158 * 3.58701e-8, 2)` sqmi), ~20,000 metered parking spots (`r round(20000 * 158 * 3.58701e-8, 2)` sqmi), and ~230,000 "other" street spaces (`r round(230000 * 158 * 3.58701e-8, 2)` sqmi)
* the area of DC's single-family home parking spaces, assuming one parking space per single-family house (`r round((39451 + 78711) * 158 * 3.58701e-8, 2)` sqmi)

```{r}
pk_area <- drop_units(sf::st_area(sf::st_union(pk, by_feature = F))) * 3.86102e-7

# https://ggwash.org/view/90490/street-parking-in-dc-is-way-too-cheap
# https://www.point2homes.com/US/Neighborhood/DC/Washington-Demographics.html
ps_area <-
  pk_area +                      # parking lots from open data dc
  (150000 * 158 * 3.58701e-8) +  # RPP spots
  (20000 * 158 * 3.58701e-8) +   # metered spots
  (230000 * 158 * 3.58701e-8) +  # "other" spaces
  ((39451 + 78711) * 158 * 3.58701e-8)  # SFH spots

parking_spots_poly <- make_polygon(center_lat, center_lon, sqmi = ps_area)

p <- ggplot() +
  
  # Basemap raster, masked to DC boundary
  geom_spatraster_rgb(data = basemap_masked) +
  
  # DC boundary outline
  geom_sf(data      = dc_boundary,
          fill      = NA,
          color     = "grey30",
          linewidth = 0.6) +
  
  # Capitol Hill square polygon
  geom_sf(data      = parking_spots_poly,
          fill      = "#B22222",
          color     = "#B22222",
          alpha     = 0.25,
          linewidth = 0.8) +
  
  # Crop plot to DC bounding box
  coord_sf(
    xlim   = st_bbox(dc_boundary)[c("xmin", "xmax")],
    ylim   = st_bbox(dc_boundary)[c("ymin", "ymax")],
    expand = FALSE
  ) +
  
  # North arrow + scale bar
  # annotation_north_arrow(
  #   location    = "tr",
  #   which_north = "true",
  #   style       = north_arrow_fancy_orienteering(text_size = 8)
  # ) +
  # annotation_scale(location = "bl", width_hint = 0.3) +
  
  labs(
    # title    = "Capitol Hill — Private Vehicles Coverage (~1.79 sq mi)",
    # subtitle = "Washington, D.C.",
    caption  = "Basemap CartoDB / OpenStreetMap contributors"
  ) +
  theme_void(base_size = 12) +
  theme(
    plot.title    = element_text(face = "bold", size = 13, margin = margin(b = 4)),
    plot.subtitle = element_text(color = "grey40", size = 10),
    plot.caption  = element_text(color = "grey60", size = 7,
                                 margin = margin(t = 6)),
    plot.margin   = margin(10, 10, 10, 10)
  )

p
```

This covers over `r round(ps_area)` square miles. This is still an underestimate. It doesn't include underground parking, for example, or account for the fact that many single family homes have 2 parking spaces.


```{r leaflet_map, eval=F, include=F, echo=F}

# 
# make_polygon <- function(center_lat, center_lon, sqmi) {
#   # 1.8 square miles = 1.8 * 2589988.11 square meters
#   target_area_m2 <- sqmi * 2589988.11
#   
#   # Calculate side length of a square with that area (in meters)
#   side_m <- sqrt(target_area_m2)
#   
#   # Convert side length to degrees
#   # 1 degree latitude ~ 111,320 meters
#   # 1 degree longitude ~ 111,320 * cos(lat) meters
#   lat_deg <- side_m / 111320
#   lon_deg <- side_m / (111320 * cos(center_lat * pi / 180))
#   
#   # Build the polygon corners
#   half_lat <- lat_deg / 2
#   half_lon <- lon_deg / 2
#   
#   polygon_coords <- matrix(
#     c(
#       center_lon - half_lon, center_lat - half_lat,
#       center_lon + half_lon, center_lat - half_lat,
#       center_lon + half_lon, center_lat + half_lat,
#       center_lon - half_lon, center_lat + half_lat,
#       center_lon - half_lon, center_lat - half_lat  # close the polygon
#     ),
#     ncol = 2, byrow = TRUE
#   )
#   
#   # Create an sf polygon
#   return(
#     st_sfc(
#       st_polygon(list(polygon_coords)),
#       crs = 4326
#     )
#   ) 
# }
# 
# # Capitol Hill neighborhood center (approximate)
# center_lat <- 38.8897
# center_lon <- -76.993
# 
# private_vehicles_poly <- make_polygon(center_lat, center_lon, sqmi = 1.79)
# # all_vehicles_poly <- make_polygon(center_lat, center_lon, sqmi = 2)
# 
# # Draw the map
# leaflet() |>
#   addProviderTiles(providers$CartoDB.Positron) |>
#   # addPolygons(
#   #   data       = all_vehicles_poly,
#   #   color      = "orange",
#   #   weight     = 2,
#   #   opacity    = 1,
#   #   fillColor  = "orange",
#   #   fillOpacity = 0.2,
#   #   label      = "All vehicles (~2 sq mi)"
#   # ) |>
#   addPolygons(
#     data       = private_vehicles_poly,
#     color      = "#B22222",
#     weight     = 2,
#     opacity    = 1,
#     fillColor  = "#B22222",
#     fillOpacity = 0.2,
#     label      = "All private vehicles (~1.79 sq mi)"
#   ) |>
#   setView(lng = center_lon, lat = center_lat, zoom = 12)
```


```{r}
# table(co$`Count of Vehicles`)
```

This write up will analyze the distribution of vehicle ownership in the data set DMV provided.
<br>
<br>


```{r}
# round(prop.table(table(co$`Count of Vehicles`))*100, 5)
```


```{r}
pcts <- prop.table(table(co$`Count of Vehicles`))*100

car_stats <- as.numeric(names(pcts)) * pcts

w_pcts <- car_stats / sum(car_stats) * 100
w_pcts_trunc <- round(c(w_pcts[1:3], "4" = sum(w_pcts[4:length(w_pcts)])))
pcts_trunc <- round(c(pcts[1:3], "4" = sum(pcts[4:length(pcts)])))





df1 <-
  data.frame(
    c(rep(1, pcts_trunc["1"]),
      rep(2, pcts_trunc["2"]),
      rep(3, pcts_trunc["3"]),
      rep(4, pcts_trunc["4"])
      )
  )
names(df1) <- "cc"
df1$ones <- 1
df1$cc <- as.factor(df1$cc)
# due to rounding we have one to many boxes (101), so remove the last box, since
# the 4+ group was farthest from being rounded up (even though it was)
df1 <- df1[1:100,]  


df2 <-
  data.frame(
    c(rep(1, w_pcts_trunc["1"]),
      rep(2, w_pcts_trunc["2"]),
      rep(3, w_pcts_trunc["3"]),
      rep(4, w_pcts_trunc["4"])
      )
  )
names(df2) <- "cc"
df2$ones <- 1
df2$cc <- as.factor(df2$cc)


waffle_data <- waffle_iron(df1, aes_d(group = cc), rows = 4)

p1 <-
  ggplot(waffle_data, aes(x, y, fill = group)) + 
  geom_waffle() + 
  coord_equal() + 
  scale_fill_manual(
                    name = "# cars reg. at address",
                    values = c("#ABEDC4", "#35D472", "#19763D", "#072211"),
                    labels = c("1", "2", "3", "4+")
                      ) +
  theme_waffle() +
  theme(legend.position = "none") +
  xlab("") +
  ylab("") +
  ggtitle("For example: only about 1.6% of households own 4 or more vehicles...") +
  theme(plot.title = element_text(size = 12))


waffle_data <- waffle_iron(df2, aes_d(group = cc), rows = 4)

p2 <-
  ggplot(waffle_data, aes(x, y, fill = group)) + 
  geom_waffle() + 
  coord_equal() + 
  scale_fill_manual(
                    name = "# cars registered at address",
                    values = c("#ABEDC4", "#35D472", "#19763D", "#072211"),
                    labels = c("1", "2", "3", "4+")
                      ) +
  theme_waffle() +
  theme(legend.position = "bottom") +
  xlab("") +
  ylab("") +
  ggtitle("But those households account for almost 6% of vehicles registered in DC") +
  theme(plot.title = element_text(size = 12))


p1 / p2 + 
  plot_annotation('Households with multiple vehicles account for disproportionate share of total vehicles', 
                  caption = paste0(
                    "There are 100 squares in each plot.\n", 
                    "In the first plot, each square represents roughly 900 addresses/households. (There are 89,871 total addresses in the data set).\n",
                    "In the second plot, each square represents roughly 1,200 vehicles. (There are 118,907 total vehicles in the data set)."),
                          theme=theme(plot.title=element_text(hjust=0.5),
                                      plot.caption = element_text(hjust = 0))
                  )

```

According to this FOIA'd data, here is the number of vehicles registered in each ANC Single Member District (SMD):

```{r}
smd_co <- 
  sf::st_join(x = smd, y = co) %>%
  sf::st_drop_geometry() %>%
  group_by(SMD_ID) %>%
  summarize(nv = sum(`Count of Vehicles`, na.rm=T))

smd_co <- dplyr::left_join(smd %>% select(SMD_ID), smd_co, by="SMD_ID")


# pal <- colorNumeric(
#   palette = "YlOrRd", 
#   domain = smd_co$nv
# )
# 
# leaflet() %>% 
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   addPolygons(
#   data = smd_co, 
#   fillColor = ~pal(nv),      # Fill color of the polygons
#   color = "grey",          # Border color
#   weight = 1,              # Border thickness
#   fillOpacity = 0.6,       # Fill opacity
#   label = ~paste0("SMD ", SMD_ID, ": ", `nv`)            # Add interactive popups (e.g., county name)
#   )

ggplot(smd_co) +
  geom_sf(aes(fill = nv), color = "grey", linewidth = 0.3) +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "YlOrRd"),
    name = "Registered\nVehicles"
  ) +
  labs(
    title = "Registered Vehicles by ANC Single Member District"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )

```



Here is the number of vehicles registered as a percentage of the number of households in the tract in 2020.

```{r}
trt_co <- 
  sf::st_join(x = trt %>% sf::st_transform(4326), y = co) %>%
  sf::st_drop_geometry() %>%
  group_by(TRACT) %>%
  summarize(
            nv = sum(`Count of Vehicles`, na.rm=T),
            num_hh = max(num_hh, na.rm=T), 
            pop_over_18 = max(pop_over_18, na.rm=T)
            ) %>%
  mutate(
          nv_hh = round(nv / num_hh*100),
          nv_po = round(nv / pop_over_18*100)
          ) %>%
  mutate(
    nv_hh = ifelse(nv_hh > 100, 100, nv_hh),
    nv_po = ifelse(nv_po > 100, 100, nv_po)
  )

trt_co <- dplyr::left_join(trt %>% select(TRACT), trt_co, by="TRACT")

# pal <- colorNumeric(
#   palette = "YlOrRd", 
#   domain = trt_co$nv_hh
# )
# 
# leaflet() %>% 
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   addPolygons(
#   data = trt_co %>% sf::st_transform(4326), 
#   fillColor = ~pal(nv_hh),      # Fill color of the polygons
#   color = "grey",          # Border color
#   weight = 1,              # Border thickness
#   fillOpacity = 0.6,       # Fill opacity
#   label = ~paste0("Tract ", TRACT, ": ", nv_hh, "% of 2020 HH")            # Add interactive popups (e.g., county name)
#   )



ggplot(trt_co %>% sf::st_transform(4326)) +
  geom_sf(aes(fill = nv_hh), color = "grey", linewidth = 0.3) +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "YlOrRd"),
    name = "% of 2020\nHouseholds"
  ) +
  labs(
    title = "Vehicles per Household by Census Tract"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```


We can correlate median household income in each tract in the 2019-2023 5-year American Community Survey with our "# vehicles / # households" statistic. When we do this, we find that every $25,000 increase in tract median income is associated with an increase of 7 additional cars per household.


```{r}
trt_d_co <- 
  sf::st_join(x = trt_d %>% sf::st_transform(4326), y = co) %>%
  sf::st_drop_geometry() %>%
  group_by(TRACT) %>%
  summarize(
            nv = sum(`Count of Vehicles`, na.rm=T),
            med_hh_inc = max(med_hh_inc, na.rm=T)
            ) %>%
  dplyr::left_join(y = trt %>% sf::st_drop_geometry(), by = "TRACT") %>%
    mutate(
          nv_hh = round(nv / num_hh*100),
          nv_po = round(nv / pop_over_18*100)
          ) %>%
  mutate(
    nv_hh = ifelse(nv_hh > 100, 100, nv_hh),
    nv_po = ifelse(nv_po > 100, 100, nv_po)
  ) %>% 
  filter(med_hh_inc != -Inf)


# GET EQUATION AND R-SQUARED AS STRING
# SOURCE: https://groups.google.com/forum/#!topic/ggplot2/1TgH-kG5XMA

lm_eqn <- function(df, x, y){
    m <- lm( formula(paste(y, " ~ ", x)), df);
    eq <- substitute(italic(y) == a + b %.% italic(x)*","~~italic(r)^2~"="~r2, 
         list(a = format(unname(coef(m)[1]), digits = 2),
              b = format(unname(coef(m)[2]), digits = 2),
             r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq));
}

ggplot(data = trt_d_co,
       aes(x = med_hh_inc/1000, y = nv_hh)) +
  geom_point() +
  geom_smooth(method = "lm") +
  geom_text(x = 50000, y = 80, label = lm_eqn(trt_d_co, x = "med_hh_inc", "nv_hh"), parse = TRUE) +
  ylab("# vehicles / # households") +
  xlab("Median household income (thousands of $)") +
  theme_minimal() +
  ggtitle("Richer areas have more vehicles") +
  labs(caption = "Note: tract income is top-coded at $250,000") +
  theme(plot.caption = element_text(hjust = 0))
```


```{r}
mto_buffer <- sf::st_buffer(x = mto, dist = 800) %>% sf::st_transform(4326)
n_cars_by_metro <-
  sf::st_join(x = sf::st_as_sf(sf::st_union(mto_buffer, by_feature = F)), y = co) %>%
  sf::st_drop_geometry() %>%
  # group_by(NAME) %>%
  summarize(nv = sum(`Count of Vehicles`, na.rm=T)) %>%
  pull(nv)

n_cars_by_metro_d <-
  sf::st_join(x = mto_buffer, y = co) %>%
  # sf::st_drop_geometry() %>%
  group_by(NAME) %>%
  summarize(nv = sum(`Count of Vehicles`, na.rm=T)) 
```

There are `r format(n_cars_by_metro, big.mark=",")` registered vehicles within 800 meters of a metro station in DC according to this data set. That is `r round(n_cars_by_metro / sum(co[["Count of Vehicles"]]) * 100)`% of the total number of vehicles.

```{r, echo=F, warning=F, message=F, include=F}
# n_cars_pal <- colorNumeric(palette = "YlOrRd", domain = n_cars_by_metro_d$nv)
# 
# leaflet() %>% 
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   addPolygons(
#   data = n_cars_by_metro_d, 
#   fillColor = ~n_cars_pal(nv),      # Fill color of the polygons
#   weight = 0,              # Border thickness
#   fillOpacity = 0.6        # Fill opacity
#   )

# Get DC boundary
dc_boundary <- counties(state = "DC", cb = TRUE) %>%
  sf::st_transform(sf::st_crs(n_cars_by_metro_d))

# Download and mask basemap to DC boundary
basemap <- get_tiles(
  dc_boundary,
  provider = "CartoDB.Positron",
  zoom = 13,
  crop = TRUE
)
basemap_masked <- terra::mask(basemap, terra::vect(dc_boundary))
```

```{r}
# Plot
ggplot() +
  geom_spatraster_rgb(data = basemap_masked) +
  geom_sf(
    data = n_cars_by_metro_d,
    aes(fill = nv),
    color = NA,
    alpha = 0.8
  ) +
  geom_sf(
    data = dc_boundary,
    fill = NA,
    color = "black",
    linewidth = 0.5
  ) +
  scale_fill_gradientn(
    colors = RColorBrewer::brewer.pal(9, "YlOrRd"),
    name = "Registered\nVehicles"
  ) +
  coord_sf(
    xlim = sf::st_bbox(dc_boundary)[c("xmin", "xmax")],
    ylim = sf::st_bbox(dc_boundary)[c("ymin", "ymax")]
  ) +
  labs(title = "There are over 43,000 vehicles registered\nwithin a 10-minute walk of a metro station") +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 9)
  )
```


```{r}
# leaflet() %>% 
#   addProviderTiles(providers$CartoDB.Positron) %>%
#   addPolygons(
#   data = sf::st_union(mto_buffer, by_feature = F), 
#   fillColor = "#487441",      # Fill color of the polygons
#   color = "grey",          # Border color
#   weight = 1,              # Border thickness
#   fillOpacity = 0.6        # Fill opacity
#   )
# 
# # Get DC boundary
# dc_boundary <- counties(state = "DC", cb = TRUE) %>%
#   sf::st_transform(sf::st_crs(mto_buffer))
# 
# # Get the buffer union
# buffer_union <- sf::st_union(mto_buffer, by_feature = F)
# 
# # Download basemap tiles cropped to DC extent
# basemap <- get_tiles(
#   dc_boundary,
#   provider = "CartoDB.Positron",
#   zoom = 13,
#   crop = TRUE
# )
# 
# # Mask the basemap raster to the DC boundary
# dc_boundary_vect <- terra::vect(dc_boundary)
# basemap_masked <- terra::mask(basemap, dc_boundary_vect)
# 
# # Plot
# ggplot() +
#   geom_spatraster_rgb(data = basemap_masked) +
#   geom_sf(
#     data = buffer_union,
#     fill = "#487441",
#     color = "grey",
#     linewidth = 0.3,
#     alpha = 0.6
#   ) +
#   geom_sf(
#     data = dc_boundary,
#     fill = NA,
#     color = "black",
#     linewidth = 0.5
#   ) +
#   coord_sf(
#     xlim = sf::st_bbox(dc_boundary)[c("xmin", "xmax")],
#     ylim = sf::st_bbox(dc_boundary)[c("ymin", "ymax")]
#   ) +
#   theme_void()
```



The map and table below shows the data at a granular level.


```{r}

co_jittered <- co

min_j <- 0.00001
co_jittered$lon <- st_coordinates(co_jittered)[, 1] + runif(n = nrow(co), min = min_j, max=min_j*10)
co_jittered$lat <- st_coordinates(co_jittered)[, 2] + runif(n = nrow(co), min = min_j, max=min_j*10)

leaflet(co_jittered) %>%
  addTiles(
    options = providerTileOptions(
      # maxNativeZoom = 19, # The maximum zoom level provided by the tile source
      maxZoom = 22        # The maximum zoom level allowed for the user
    )
  ) %>%  # Add default OpenStreetMap tiles
  addCircleMarkers(
    lng = ~lon,
    lat = ~lat,
    radius = ~`Count of Vehicles` + 2,
    color = "purple",
    fillOpacity = .7,
    stroke=F,
    label = ~paste0(`Count of Vehicles`, ": ", Address),
    group="vehicles",
    clusterOptions = markerClusterOptions(
      # Markers will not be clustered at zoom level 10 and below (more zoomed in)
      disableClusteringAtZoom = 16
    )
      )

```

```{r, warning=F, message=F}
co %>% sf::st_drop_geometry() %>% 
  select(Address, `Count of Vehicles`) %>% 
  # slice(1:10000) %>%
  DT::datatable(
  .,
  filter = 'top', # or 'bottom'
  options = list(pageLength = 10) # Optional: set default number of entries shown
)
```

